---
title: "Chapter_3_R_for_Data_Science"
author: "Trevor Barnes"
date: "11/18/2017"
output: html_document
runtime: shiny
---
# Chapter 3 - Data Transformation with dplyr

## nycflights13 - be sure that you have run install.packages(nycflights13)
```{r, echo=FALSE}
library(nycflights13)
library(tidyverse)

flights
```

## Filter Rows with filter()

```{r, error=TRUE}

## Filter for month and day being 1,1

filter(flights, month ==1, day ==1)

## Practice setting variables to the filters

Jan1 <- filter(flights, month == 1, day == 1)
(Dec25 <- filter(flights, month == 12, day == 25))

## Be sure to use the '==' and not just '='. See how the following results in an error

filter(flights, month = 1)

## Another error you may receive happens with floating point numbers

sqrt(2) ^ 2 == 2

1/49 * 49 == 1

## Try using the near function in these instances

near(sqrt(2)^2, 2)

near(1/49*49, 1)
```

## Logical Operators - & is 'and' - | is 'or' - ! is 'not'

```{r}

## Filter flights for Nov or Dec

filter(flights, month == 11 | month == 12)

## Retyping the same field over and over can be tedious
## Try using the 'x %in% y' which will select every row where x is the value of y

nov_dec <- filter(flights, month %in% c(11,12))

## De Morgan's law states that !(x & y) is the same as !x | !y
## Also, that !(x | y) is the same as !x & !y
## The following formula can be written two different ways to find flights that were not delayed on departure or arrival by more than two hours

filter(flights, !(arr_delay > 120 | dep_delay > 120))

filter(flights, arr_delay <= 120, dep_delay <= 120)
```

## Missing Values

```{r, error=TRUE}

## Missing Values (NA) can be an issue in R, they represent unknown value so they become contagious, infecting the function resulting in an unknown output

NA > 5

10 == NA

NA + 10

NA / 2

## The most confusing being

NA == NA

## To reduce confusion, lets pretend x is Mary's age and y is John's age and we don't know how old they are

x <- NA

y <- NA

x == y

## If you want to verify if a value is missing, use the 'is.na' function

is.na(x)

## Filter only includes rows where the condition is TRUE, it excldes both FALSE and NA

df <- tibble(x = c(1, NA, 3))

filter(df, x > 1)

filter(df, is.na(x) | x > 1)
```

## Arrange Rows with arrange()

```{r}

## Example

arrange(flights, year, month, day)

## You can use 'desc()' to sort a column in descending order

arrange(flights, desc(arr_delay))

## Missing values are always sorted at the end

df <- tibble(x = c(5,2,NA))

arrange(df, x)

arrange(df, desc(x))
```

## Select Columns with select()

```{r}

## This is very useful when working with many columns

select(flights, year, month, day)

## Select all columns between two columns

select(flights, year:day)

## Select all columns except from between those columns

select(flights, - (year:day))

## Some other helpful select 
## starts_with("abc")
## ends_with("xyz")
## contains("ijk")
## matches("(.)\\1")
## num_range("x", 1:3) ## matches x1, x2, x3
## see ?select for more details

## select can also be used to rename variables, as 'rename()' is a version of select

rename(flights, tail_num = tailnum)

## Another option is to use 'select()' in conjunction with the 'everything()'hepler and is useful if you have variables you want to move at the start of the data frame

select(flights, time_hour, air_time, everything())
```

## Add new variables with mutate()

```{r}

## Mutations work by adding new columns at the end of your dataset, it may be easier to first narrow down your dataset

flights_sml <- select(flights, 
                      year:day,
                      ends_with("delay"),
                      distance,
                      air_time
                      )

mutate(flights_sml,
       gain = arr_delay - dep_delay,
       speed = distance / air_time * 60)

## Note that you can refer to columns that you just created

mutate(flights_sml,
       gain = arr_delay - dep_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours
       )

## If you only want to keep the new variables than use transmute()

transmute(flights,
          gain = arr_delay - dep_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours)
```

## Useful Creation Functions

```{r}

## By now you should be familiar with the +, -, *, /, and ^ arithmatic operators

## But modular arithematic operators are very powerful, too
## '%/%' is integer division and '%%' is remainder

transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100
          )

## Use 'log()', 'log2()', and 'log10()' to transform data ranging across multiple magnitudes

## Use 'lead()' and 'lag()' that helps compute running differences

(x <- 1:10)

lag(x)

lead(x)

## Compute cumulative and rolling aggregates with the different cum- functions
## 'sumsum()', 'cumprod()', 'cummin()', cummax()', 'cummean()'

x

cumsum(x)

cummin(x)

cummean(x)

## Logical comparisons include <, <=, >, >=, !=

## Ranking is also another powerful tool in sorting data

y <- c(1, 2, 2, NA, 3, 4)

min_rank(y)

min_rank(desc(y))

## If 'min_rank()' doesnt do what you want it to do, look at the variants with these functions
## 'row_number()', 'dense_rank()', 'percent_rank()', 'cume_dist()', 'ntile()'

row_number(y)

dense_rank(y)

percent_rank(y)

cume_dist(y)
```

## Grouped Summaries with summarize()

```{r}

## This function will help summarize data by collapsing it to a single row

summarize(flights, delay = mean(dep_delay, na.rm = TRUE))

## Summzarize doesn't become very useful until we pair it with a group by

by_day <- group_by(flights, year, month, day)

summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

## Combining Multiple Operations with Piping

```{r}

## Piping is  avery useful tool for preventing rewriting code over and over within a function, for example, look how we would normally script the following to summarize delays (HNL is an outlier so we remove it)

by_dest <- group_by(flights, dest)

delay <- summarize(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE)
                   )
delay <- filter(delay, count > 20, dest != "HNL")


## Notice above how there are 3 different steps to prepare the data
##    1) Group the flights by destination
##    2) Summarize to compute average distance, average delay, and number of fights
##    3) Filter to remove noisy points around Honolulu, which is twice as far, on average than          the other airports

## Let us view the data in a plot

ggplot(data = delay,
       mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

## Let us try to make the code more effecient by 'piping' it using %>%

delays <- flights %>%
  group_by(dest) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dest != "HNL")

## Why do we keep using the 'na.rm = TRUE' argument? Let us take a look at our data without it

flights %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay))

## We get a lot of missing values, lets add back in the 'na.rm = TRUE'

flights %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay), na.rm = TRUE)

## We can attack this from a different angle and first remove the flights that have been cancelled

not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay))
```

## Counts

```{r}

## When you are doing data aggregation, it is always good to do a count, 'n()' or count the nonmissing values, '!is.na(x)'. This way you can verify that you are not drawing conclusions based on small amounts of data

delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)

## We can see that there are planes that have ana average delay of 300 minutes (5 hours), but this isn't giving us the whole picture. We can get a better look if we draw a scatterplot

delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n=n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)

## It can be helpful to filter out data with very few observations, let us now remove tailnums with less than 25 observations

delays %>%
  filter(n>25) %>%
  ggplot(mapping = aes(x = n, y = delay)) +
    geom_point(alpha = 1/10)

## Let us practice with a different set of data, first, convert our data to a tibble so it looks nice

batting <- as.tibble(Lahman::Batting)

batters <- batting %>%
  group_by(playerID) %>%
  summarize(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>%
  filter(ab > 100) %>%
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() +
    geom_smooth(se = FALSE)

## We can also exaplin some of the outlying data by arranging, in this case, we can explain away people who have extermely high batting averages as people who were lucky, not skilled

batters %>%
  arrange(desc(ba))
```

## Useful Summary Functions

```{r}

## Measures of Location include the 'mean()' and 'median()' functions, which helps in combining aggregation and subsetting the data

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    ## Average Delay
    avg_delay1 = mean(arr_delay),
    ## Average Positive Delay (Removes 0's)
    avg_delay2 = mean(arr_delay[arr_delay > 0])
  )

## Measures of Spread include the 'sd()', 'IQR()', and 'mad()' fucntions and are useful if you have outliers

not_cancelled %>%
  group_by(dest) %>%
  summarize(distance_sd = sd(distance)) %>%
  arrange(desc(distance_sd))


## Measures of Rank include the 'min()', 'quantile()', and 'max()' functions

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    first = min(dep_time),
    last = max(dep_time)
  )

## Measures of Position include the 'first()', 'ntn()', and 'last()' functions

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    first_dep = first(dep_time),
    last_dep = last(dep_time)
  )

## These functions are complimentary to filtering on ranks. Filtering gives you all variables with each observation in a seperate row

not_cancelled %>%
  group_by(year, month, day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))

## Counts are useful for taking no arguments, but returning the total count of a group like 'n()', or to count non-missing values use 'is.na()'. To count unique (distinct) values use 'n_distinct()'

not_cancelled %>%
  group_by(dest) %>%
  summarize(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))

## If you just want a simple count

not_cancelled %>%
  count(dest)

## You can additionally provide a weight variable

not_cancelled %>%
  count(tailnum, wt=distance)

## Counts and proportions of logical values when used with numeric functions convert True to 1 and False to 0. This makes the 'sum()' and 'mean()' value very useful

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(n_early = sum(dep_time < 500))

## What proportion of flights are delayed by more than an hour

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(hour_perc = mean(arr_delay > 60))

## Grouping by multiple variables can help you easily peel off variables that makes it progressively earlier to roll up a dataset

daily <- group_by(flights, year, month, day)

per_day <- summarize(daily, flights = n())

per_month <- summarize(per_day, flights = sum(flights))

per_year <- summarize(per_month, flights = sum(flights))

## Ungrouping can be helpful, in this case to no longer group flights by date

daily %>%
  ungroup() %>%
  summarize(flights=n())
```

## Grouped Mutates

```{r}

## Grouping is most useful in ocnjunction with 'summarize()', but you can also do convenient operations with 'mutate()' and 'filter()'

flights_sml %>%
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)

## Filter all groups bigger than a threshold

popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n() > 365)

## Standardize to compute per group metrics

popular_dests %>%
  filter(arr_delay > 0) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(year:dest, dest, arr_delay, prop_delay)
```