---
title: 'Chapter 12: Factor with forcats'
author: "Trevor Barnes"
date: "March 4, 2018"
output: html_document
---



```{r}
##Library Load

library(tidyverse)
library(forcats)
library(readr)
```

### Creating Factors

Imagine you have a variable that records month



```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
```

Using a String to record variables has two problems:

  1. There are only 12 possible months and there's nothing saving you from typos
  
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
```

  2. It doesn't sort in a useful way
  
You can fix both of these problems with a factor. To create a factor you must start by creating a list of valid *levels*:

```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

Now you can create a factor

```{r}
y1 <- factor(x1, levels = month_levels)

y1

sort(y1)
```

And any values not in the set will be silently converted to NA:

```{r}
y2 <- factor(x2, levels = month_levels)

y2
```

If you want to know about the error, you can use readr::parse_factor():

```{r,, error = TRUE}
y2 <- parse_factor(x2, levels = month_levels)
```

If you omit the levels they will be taken from the data in alphabetical order:

```{r}
factor(x1)
```

Sometimes you want the order of the levels to match the order they first appear in the data, use the *unique()* and after that *fct_inorder()*

```{r}
f1 <- factor(x1, levels = unique(x1))
f1

f2 <- x1 %>%
  factor() %>%
  fct_inorder()
```

If you ever need to access the set of valid levels directly, you can do so with *levels()*:

```{r}
levels(f2)
```

### General Social Survey

For the rest of this chapter we will focus on *forcats::gss_cat*. It's a sample of data from General Social Survey, which is a lon running US Survey. Call the data:

```{r}
gss_cat
```

When factors are stored in a level, it is difficult to see there levels. One way to see them is to use *count()*

```{r}
gss_cat %>%
  count(race)
```

Or with a bar chart:

```{r}
ggplot(gss_cat,aes(race)) +
  geom_bar()
```

By default ggplot2 will drop levels that don't have any values. You can force them to display with:

```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

These levels represent valid values that simply did not occur in this dataset. When working with factors the two most common operations are:

  1. Changing the *order* of levels
  2. Changing the *values* of levels

### Modifying Factor Order

It is often useful to change the order of the factor levels in a visualization. For example, imagine you want to explore the average number of hours spent watching TV per day across religions:

```{r}
relig <- gss_cat %>%
  group_by(relig) %>%
  summarize(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n=n()
  )

ggplot(relig, aes(tvhours, relig)) +
  geom_point()
```

IT is difficult to interpret this plot because there is no overall pattern. We can improve it by reordering the levels of *relig* using *fct_reorder()*. *fct_reorder()* takes three arguments:
  
  * f, the factor whose levels we want to modify
  * x, a numeric vectorthat you want to use to reorder the levels
  * Optionally, fun, a function that's used if there are multiple value of x for each value of f. the default value is median
  
Reordering religion makes it much easier to see that people in the "Don't Know" category watch much more TV, and Hinduism and other Eastern Religions watch much less. As you start making more complicated Transformations try using *aes()* and into a separate *mutate()* step:

```{r}
relig %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours,relig)) +
    geom_point()
```

What if we create a similar plot looking at how average age varies across reported income level?

```{r}
rincome <- gss_cat %>%
  group_by(rincome) %>%
  summarize(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n=n()
  )

ggplot(
  rincome,
  aes(age, fct_reorder(rincome,age))
) + geom_point()
```

Reordering the levels here isn't a good idea, because income was already ordered. It does make sense to pull "Not Applicable" to the front with the other special levels. You can use *fct_relevel()*. It takes a factor ,f, and thean any number of levels that you weant o move to the front of the line: 

```{r}
ggplot(
  rincome,
  aes(age, fct_relevel(rincome,"Not Applicable"))
) + 
  geom_point()
```

Another type of reordering is useful when you are coloring the lines on a plot. *fct_reorder2()* reorders the factor by the y values associated with the largest x values. This makes the plot easier to read because of the line colors line up with the legend:

```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  group_by(age, marital) %>%
  count()%>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, color = marital)) +
  geom_line(na.rm = TRUE)

ggplot(
  by_age,
  aes(age, prop, color = fct_reorder2(marital, age, prop))
) +
  geom_line() +
  labs(color = "marital")
```

Finally, for bar plots, you can use *fct_infreq()* to order levels in increasing frequency: thjis is the simplest type of reorderingbecause it doesn't need any extra variables. You may want to combine with *fct_rev()*:

```{r}
gss_cat%>%
  mutate(marital = marital%>%
           fct_infreq()%>%
           fct_rev())%>%
  ggplot(aes(marital)) +
    geom_bar()
```

### Modifying Factor Levels

More powerful than changing the orders of levels is to change their values. This allows you to clarify labels for publication, and collapse levels for high-level displays. The most general and powerful tool is *fct_recode()*. It allows you to recode, or change, the value at each level. For example, take gss_cat$partyid:

```{r}
gss_cat %>%
  count(partyid)
```

The levels are terse and inconsistent. Let's tweak them to be longer ans use a parallel construction:

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong"    = "Strong republican",
                              "Republican, weak"      = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak"        = "Not str democrat",
                              "Democrat, strong"      = "Strong democrat"
                              )) %>%
  count(partyid)
```

*fct_recod()* will leave levels that aren't texplicitly mentioned as is, and will warn you if you accidentally refer to a level that doesn't exist. To combine groups, you can assign multiple old levels to the same new level.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong"    = "Strong republican",
                              "Republican, weak"      = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak"        = "Not str democrat",
                              "Democrat, strong"      = "Strong democrat",
                              "Other"                 = "No Answer",
                              "Other"                 = "Don't Know",
                              "Other"                 = "Other party"
                              )) %>%
  count(partyid)
```

You must use this technique with care: if you group together categories that are truly different you will end up with misleading results.

If you want to collapse a lot of levels, *fct_collapse()* is a useful variant of *fct_recode()*. For each new variable, you can provide a vector of old levels:

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c("No Answer", "Don't know", "Other party"),
                                rep = c("Strong republican", "Not str republican"),
                                ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                dem = c("Not str democrat", "Strong democrat")
                                )) %>%
  count(partyid)
```

Sometimes you just want to lump together all the small groups to make a plot or table simpler. That is the job of *fct_lump()*:

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
```

The default behaviour is to progressively lump together the smallest groups, ensuring that the aggregate si still the smallest group. In this case it's not very helpful: it is true that the majority of Americans in this survey are Protestant, but we've probably overcollapsed.

Instead, we can use the n parameter to specify how many groups (excluding other) we want to keep:

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n =10)) %>%
  count(relig, sort = TRUE) %>%
  print(n =Inf)
```